Counterfactual Training: Teaching Models Plausible and Actionable Explanations

### 2025-07-22 paper/preprint/paper.tex

\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-train-default}: Training Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-eval-default}: Evaluation Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]

@@

\subsection{Confidence Intervals}\label{confidence-intervals}

@@

\subsection{Qualitative Findings for Image
Data}\label{qualitative-findings-for-image-data}

@@

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-tune-train}: Training Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]

>>>>

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-train-default}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-eval-default}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]

@@

\subsubsection{Confidence Intervals}\label{confidence-intervals}

@@

\subsubsection{Qualitative Findings for Image
Data}\label{qualitative-findings-for-image-data}

@@

\subsubsection{Costs}\label{costs}

\caption{\label{tbl-costs}Costs}

\begin{tabular}{lc}
  \toprule
  \textbf{Data} & \textbf{Cost $(-\%)$} \\\midrule
  LS & -26.82\pm0.86 $^{*}$ \\
  Circ & 40.97\pm0.82 $^{*}$ \\
  Moon & 33.83\pm0.98 $^{*}$ \\
  OL & 10.35\pm1.28 $^{*}$ \\\midrule
  Adult & 1.16\pm3.53 $^{}$ \\
  CH & -34.89\pm2.31 $^{*}$ \\
  Cred & 28.24\pm1.08 $^{*}$ \\
  GMSC & 3.54\pm5.78 $^{}$ \\
  MNIST & -31.67\pm7.72 $^{*}$ \\\midrule
  Avg. & 2.75 \\\bottomrule
\end{tabular}

\subsubsection{Validity}\label{validity}

\caption{\label{tbl-val}Validity}

\begin{tabular}{lcc}
  \toprule
  \textbf{Data} & \textbf{CT} & \textbf{BL} \\\midrule
  LS & 1.0 & 1.0 \\
  Circ & 0.97 & 0.52 \\
  Moon & 1.0 & 1.0 \\
  OL & 0.87 & 0.98 \\\midrule
  Adult & 0.61 & 0.99 \\
  CH & 0.96 & 1.0 \\
  Cred & 0.7 & 1.0 \\
  GMSC & 0.63 & 1.0 \\
  MNIST & 1.0 & 1.0 \\
\end{tabular}

\caption{\label{tbl-val-mtbl}Validity}

\begin{tabular}{lcc}
  \toprule
  \textbf{Data} & \textbf{CT} & \textbf{BL} \\\midrule
  LS & 1.0 & 1.0 \\
  Circ & 0.67 & 0.49 \\
  Moon & 0.99 & 0.98 \\
  OL & 0.37 & 0.57 \\\midrule
  Adult & 0.56 & 0.99 \\
  CH & 0.96 & 1.0 \\
  Cred & 0.67 & 1.0 \\
  GMSC & 0.38 & 1.0 \\
  MNIST & 1.0 & 1.0 \\
\end{tabular}

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-tune-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
